# III: What are hierarchical GAMs?

```{r part_3_premble, include=FALSE}
#### Code for III: What are hierarchical GAMs? ####
```

## What do we mean by hierarchical smoothers?

In this section, we will describe how to model inter-group variability using smooth curves and how to fit these models using **mgcv**. Model structure is key in this framework, so we start with three model choices:

1. Should each group have its own smoother, or will a common smoother suffice?
2. Do all of the group-specific smoothers have the same wiggliness, or should each group have its own smoothing parameter?
3. Will the smoothers for each group have a similar shape to one another --- a shared global smoother?

These three choices result in five possible models (figure \ref{fig:models}):

1. A single common smoother for all observations.
2. A global smoother plus group-level smoothers that have the same wiggliness.
3. A global smoother plus group-level smoothers with differing wiggliness.
4. Group-specific smoothers without a global smoother, but with all smoothers having the same wiggliness.
5. Group-specific smoothers with different wiggliness.

![\label{fig:models}Alternate types of functional variation f(x) that can be fitted with HGAMs. The dashed line indicates the average function value for all groups, and each solid line indicates the functional value at a given predictor value for an  individual group level. The null model (of no functional relationship between the covariate and outcome, top right), is not explicitly assigned a model number.](../figures/alternate_models.png)

It is important to note that "similar wiggliness" and "similar shape" are two distinct concepts; functions can have very similar wiggliness but very different shapes. Wiggliness measures how quickly a function changes across its range, and it is easy to construct two functions that differ in shape but have the same wiggliness. For this paper, we consider two functions to have similar shape if the average squared distance between the functions is small (assuming the functions have been scaled to have a mean value of zero across their ranges). This definition is somewhat restricted; for instance, a cyclic function would not be considered to have the same shape as a phase-shifted version of the same function, nor would two normal distributions with the same mean but different standard deviations. The benefit of this definition of shape, however, is that it is straightforward to translate into penalties akin to those described in section II. Figure \ref{fig:models}, model 4 illustrates the case where models have different shapes. Similarly, two curves could have very similar overall shape, but differ in their wiggliness. For instance, one function could be equal to another plus a high-frequency oscillation term. Figure \ref{fig:models}, model 3 illustrates this.

We will discuss the trade-offs between different models and guidelines about when each of these models is appropriate in section V. The remainder of this section will focus on how to specify each of these five models using **mgcv**.

## Coding hierarchical GAMs in R

Each of the models in Figure \ref{fig:models} can be coded straightforwardly in **mgcv**.
We will use two example datasets to demonstrate how to code these models (see the supplemental code to reproduce these examples):

A. The `CO2` dataset, available in R via the **datasets** package. This data is from an experimental study by @potvin_statistical_1990 of CO~2~ uptake in grasses under varying concentrations of CO~2~, measuring how concentration-uptake functions varied between plants from two locations (Mississippi and Quebec) and two temperature treatments (chilled and warm). Twelve plants were used and CO~2~ uptake measured at 7 CO~2~ concentrations for each plant (figure \ref{fig:vis_data}a). Here we will focus on how to use HGAMs to estimate inter-plant variation in functional responses. This data set has been modified from the default version available with R, to recode the `Plant` variable as an unordered factor `Plant_uo`[^ordered].

B. Data generated from a hypothetical study of bird movement along a migration corridor, sampled throughout the year (see supplemental code). This dataset consists of simulated sample records of numbers of observed locations of 100 tagged individuals each from six species of bird, at ten locations along a latitudinal gradient, with one observation taken every four weeks. Counts were simulated randomly for each species in each location and week by creating a species-specific migration curve that gave the probability of finding an individual of a given species in a given location, then simulated the distribution of individuals across sites using a multinomial distribution, and subsampling that using a binomial distribution to simulation observation error (i.e. not every bird present at a location would be detected). 
The data set (`bird_move`) consists of the variables `count`, `latitude`, `week` and `species` (figure \ref{fig:vis_data}b). This example allows us to demonstrate how to fit these models with interactions and with non-normal (count) data. The true model used to generate this data was model 2: a single global function plus species-specific deviations around that global function.

[^ordered]: Note that **mgcv** requires that grouping or categorical variables be coded as factors in R; it will raise an error message if passed data coded as characters. It is also important to know whether the factor is coded as ordered or unordered (see `?factor` for more details on this). This matters when fitting groupwise smoothers using the `by=` argument (as is used for fitting models 3 and 5, shown below). If the factor is unordered, **mgcv** will set up a model with one smoother for each grouping level. If the factor is ordered, **mgcv** will set any basis functions for the first grouping level to zero. In model 3 the ungrouped smoother will then correspond to the first grouping level, rather than the average functional response, and the group-specific smoothers will correspond to deviations from the first group. In model 5, using an ordered factor will result in the first group not having a smoother associated with it at all.

```{r vis_data, echo=FALSE,  fig.width=8, fig.height=4, out.width="\\linewidth", fig.cap="\\label{fig:vis_data}Example data sets used throughout section III. a) Grass $\\text{CO}\\textsubscript{2}$ uptake versus $\\text{CO}\\textsubscript{2}$ concentration for 12 individual plants. Color and linetype included to distinguish individual plant trends. b) Simulated data set of bird migration, with point size corresponding to weekly counts of 6 species along a latitudinal gradient (zeros excluded for clarity). "}

#The default CO2 plant variable is ordered;
#This recodes it to an unordered factor (see main text for why).
CO2 <- transform(CO2, Plant_uo=factor(Plant, ordered=FALSE))

#Loading simulated bird movement data
bird_move <- read.csv("../data/bird_move.csv")

CO2_vis_plot <- ggplot(CO2, aes(x=conc, y=uptake, group=Plant,color=Plant, lty=Plant)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = rep(c("red","blue","black"), times =4))+
  scale_linetype_manual(values = rep(1:4, each=3))+
  guides(color="none",linetype="none")+
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})), y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))

bird_vis_plot <- ggplot(dplyr::filter(bird_move, count > 0),
                        aes(x=week, y=latitude, size=count))+
  facet_wrap(~ species) +
  geom_point() +
  scale_size(name = "Count", range = c(0.2, 3)) +
  labs(x = "Week", y = "Latitude") +
  theme(legend.position = "bottom")

plot_grid(CO2_vis_plot, bird_vis_plot, nrow=1, labels=c("a","b"),
          align = "hv", axis = "lrtb")
```

Throughout the examples we use Restricted Maximum Likelihood (REML) to estimate model coefficients and smoothing parameters. We strongly recommend using either REML or marginal likelihood (ML) rather than the default GCV criteria when fitting GAMs, for the reasons outlined in @wood_fast_2011. In each case some data processing and manipulation has been done to obtain the graphics and results below. See supplemental code for details on data processing steps.

### A single common smoother for all observations (Model 1)

We start with the simplest model we can in our framework and include many details here to ensure that readers are comfortable with the terminology and R functions we are going to use later.

For our `CO2` data set, we will model $\log_e(\texttt{uptake})$ as a function of two smoothers: a thin plate regression spline of $\log_e$-concentration, and a random effect for plant to model plant-specific intercepts. Mathematically:

$$
\log_e(\texttt{uptake}_i) = f(\log_e(\texttt{conc}_i)) + \zeta_\texttt{Plant\_uo} + \varepsilon_i
$$

where $\zeta_\texttt{Plant\_uo}$ is the random effect for plant and $\varepsilon_i$ is a Gaussian error term. Here we assume that $\log_e(\texttt{uptake}_i)$ is normally distributed.


In R we can write our model as:
```{r co2_mod1_unrun, echo=TRUE, eval=FALSE, purl=FALSE}
CO2_mod1 <- gam(log(uptake) ~ s(log(conc), k=5, bs="tp") +
                  s(Plant_uo, k=12, bs="re"),
                data=CO2, method="REML", family="gaussian")
```

This is a common GAM structure, with a single smooth term for each variable. Specifying the model is similar to specifying a GLM in R via `glm()`, with the addition of `s()` terms to include one-dimensional or isotropic multidimensional smoothers. The first argument to `s()` are the terms to be smoothed, the type of smoother to be used for the term is specified by the `bs` argument, and the maximum number of basis functions is specified by `k`. There are different defaults in **mgcv** for k, depending on the type of smoother chosen; here we use a tprs smoother (`bs="tp"`) for the concentration smoother, and set `k=5` as there are only 7 separate values of concentration measured, so the default `k=10` (for tprs) would be too high; further, setting `k=5` saves on computational time (see section V). The random effect smoother (`bs="re"`) that we used for the `Plant_uo` factor has a default `k` equal to the number of levels in the grouping variable (here, 12). We specified `k=12` just to make this connection apparent. 

```{r co2_mod1, echo=FALSE, fig.width=6, fig.height=3, dev.args=list(pointsize=10), fig.cap="\\label{fig:co2_mod1}**mgcv** plotting output for model 1 applied to the CO2 dataset. Left shows the smoother of $\\log_e$ concentration and right plot shows a quantile-quantile plot of the random effects against Gaussian quantiles, used to check the appropriateness of the normal random effect assumption. Numbers in the vertical axis of the left figure and the title of the right give the effective degrees of freedom of the smoothers."}
<<co2_mod1_unrun>>

plot(CO2_mod1, pages=1, seWithMean=TRUE)
```

Figure \ref{fig:co2_mod1} illustrates **mgcv**'s default plotting out for `CO2_mod1`: the left panel shows the estimated smoother for concentration, and the right shows a quantile-quantile plot of the estimated random effects vs Gaussian quantiles, which can be used to check our model.

Looking at the effects by term is useful, but we are often interested in fitted values or predictions from our models. Using the built in prediction functions with **mgcv**, we can estimate what the fitted function (and uncertainty around it) should look like for each level, as shown in Figure \ref{fig:co2_mod1_predict} (see supplemental code for more details on how to generate these predictions).

```{r co2_mod1_ggplot,echo=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:co2_mod1_predict} Predicted uptake function ($\\pm$ 2 s.e.) for each plant, based on model 1 (a single global function for uptake plus a individual-level random effect intercept). Model predictions are for log-uptake, but are transformed here to show the fitted function on the original scale of the data.", out.width="\\linewidth"}

# setup prediction data
CO2_mod1_pred <- with(CO2,
                      expand.grid(conc=seq(min(conc), max(conc), length=100),
                                  Plant_uo=levels(Plant_uo)))

# make the prediction, add this and a column of standard errors to the prediction
# data.frame. Predictions are on the log scale.
CO2_mod1_pred <- cbind(CO2_mod1_pred,
                       predict(CO2_mod1, CO2_mod1_pred, se.fit=TRUE, type="response"))

# make the plot. Note here the use of the exp() function to back-transform the
# predictions (which are for log-uptake) to the original scale
ggplot(data=CO2, aes(x=conc, y=uptake, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(fit - 2*se.fit), ymax=exp(fit + 2*se.fit), x=conc),
              data=CO2_mod1_pred, alpha=0.3, inherit.aes=FALSE) +
  geom_line(aes(y=exp(fit)), data=CO2_mod1_pred) +
  geom_point() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```

Examining these plots, we see that while functional responses among plants are similar, some patterns are not captured by this model.  For instance, for plant Qc2, the model clearly underestimates CO2 uptake. A model including individual differences in functional responses may better explain variation.

For our bird example, we model the count of birds as a function of location and time, including their interaction.  For this we structure the model as:

$$
\mathbb{E}(\texttt{count}_i) = \exp(f(\texttt{week}_i, \texttt{latitude}_i))
$$

where we assume that $\texttt{count}_i \sim\text{Poisson}$. For the smooth term, $f$, we employ a tensor product of `latitude` and `week`, using a thin plate regression spline (TPRS) for the marginal latitude effect, and a cyclic cubic regression spline for the marginal week effect to account for the cyclic nature of weekly effects (we expect week 1 and week 52 to have very similar values)[^cyclic], both splines had basis complexity (`k`) of 10. 

```{r bird_mod1_comments, echo=FALSE, eval=FALSE}
# Note for specifying tensor products: you can either specify bs (basis) and
# k (number of basis functions) as single values, which would assign the same
# basis and k to each marginal value, or pass them as vectors, one value for each
# distinct marginal smoother (see ?mgcv::te for details)
```



```{r bird_mod1,  outwidth="\\linewidth", dev.args = list(pointsize=10)}
bird_mod1 <- gam(count ~ te(week, latitude, bs=c("cc", "tp"), k=c(10, 10)),
                 data=bird_move, method="REML", family="poisson",
                 knots = list(week = c(0, 52)))
```

[^cyclic]:The cyclic smoother requires that the start and end points of the cyclic variable are specified, via the `knots` argument; 
the smoother will have the exact same value at the start and end. 
In the absence of a specified start and end point, `gam` will assume the end points are the smallest and largest observed levels of the covariate (see `mgcv::smooth.construct.cc.smooth.spec` for more details). 
Note that in `bird_mod1` we have specified week 0 and week 52 as the endpoints, as the first (week 1) and last weeks (week 52) of the year should not have exactly the same expected value. 



```{r bird_mod1_plot, echo=FALSE, fig.width = 4, fig.height = 4, outwidth="\\linewidth", dev.args = list(pointsize=10), fig.cap="\\label{fig:bird_mod1} Plot illustrating the average log-abundance of all bird species at each latitude for each week, with yellow colours indicating more individuals and red colours fewer."}
#mgcv gam plot for the two-dimensional tensor product smoother for bird_mod1.
#scheme=2 displays the color scheme (rather than mgcv's default, which only
#shows contour lines)
plot(bird_mod1, pages=1, scheme=2, rug=FALSE)
box()
```

Figure \ref{fig:bird_mod1} shows the default plot (created by running `plot(bird_mod1, pages=1, scheme=2, rug=FALSE)`) for the week-by-latitude smoother. It shows birds starting at low latitudes in the winter then migrating to high latitudes from the 10th to 20th week, staying there for 15-20 weeks, then migrating back. However, the plot also indicates a large amount of variability in the timing of migration. The source of this variability is apparent when we look at the timing of migration of each species (cf. figure \ref{fig:vis_data}b).

All six species in figure \ref{fig:vis_data}b show relatively precise migration patterns, but they differ in the timing of when they leave their winter grounds and the amount of time they spend at their summer grounds. Averaging over all of this variation results in a relatively imprecise (diffuse) estimate of migration timing (figure \ref{fig:bird_mod1}), and viewing species-specific plots of observed versus predicted values (figure \ref{fig:bird-fitted-mod1}), it is apparent that the model fits some of the species better than others. This model could potentially be improved by adding inter-group variation in migration timing. The rest of this section will focus on how to model this type of variation.

```{r bird-fitted-mod1,echo=FALSE, fig.width=6, fig.height=4, out.width="\\linewidth", fig.cap="\\label{fig:bird-fitted-mod1}Observed counts by species versus predicted counts from \\texttt{bird\\_mod1} (1-1 line added as reference). If our model fitted well we would expect that all species should show similiar patterns of dispersion around the 1-1 line (and as we are assuming the data is Poisson, the variance around the mean should equal the mean). Instead we see that variance around the predicted value is much higher for species 1 and 6."}
#add the predicted values from the model to bird_move
bird_move <- transform(bird_move, mod1 = predict(bird_mod1, type="response"))

ggplot(bird_move, aes(x=mod1, y=count)) +
  facet_wrap(~species) +
  geom_point(alpha=0.1) +
  geom_abline() +
  labs(x="Predicted count", y= "Observed count")
```

### A single common smoother plus group-level smoothers that have the same wiggliness (Model 2)

Model 2 is a close analogue to a GLMM with varying slopes: all groups have similar functional responses, but inter-group variation in responses is allowed. This approach works by allowing each grouping level to have its own functional response, but penalizing functions that are too far from the average.

This can be coded in **mgcv** by explicitly specifying one term for the global smoother (as in model 1 above) then adding a second smooth term specifying the group-level smooth terms, using a penalty term that tends to draw these group-level smoothers to zero. For one-dimensional smoothers, **mgcv** provides an explicit basis type to do this, the factor-smoother interaction or `"fs"` basis (see `?mgcv::factor.smooth.interaction` for details). This smoother creates a copy of each set of basis functions for each level of the grouping variable, but only estimates one smoothing parameter for all groups. To ensure that all parts of the smoother can be shrunk towards zero effect, each component of the penalty null space is given its own penalty[^intercept_note].

[^intercept_note]: As part of the penalty construction, each group will also have its own intercept (part of the penalized null space), so there is no need to add a separate term for group specific intercepts as we did in model 1.

We modify the previous $\text{CO}_2$ model to incorporate group-level smoothers as follows:

$$
\log_e(\texttt{uptake}_i) = f(\log_e(\texttt{conc}_i)) + f_{\texttt{Plant\_uo}_i}(\log_e(\texttt{conc}_i)) + \varepsilon_i
$$

where $f_{\texttt{Plant\_uo}_i}(\log_e(\texttt{conc}_i))$ is the smoother for concentration for the given plant. In R we then have:

```{r co2_mod2, echo=TRUE}
CO2_mod2 <- gam(log(uptake) ~ s(log(conc), k=5, m=2) + 
                  s(log(conc), Plant_uo, k=5,  bs="fs", m=2),
                data=CO2, method="REML")
```

```{r co2_mod2_plot, fig.width=6, fig.height=3, dev.args=list(pointsize=10), out.width="\\linewidth", echo=FALSE, fig.cap="\\label{fig:co2_mod2}Global function (left) and group-specific deviations from the global function (right) for \\texttt{CO2\\_mod2}"}
plot(CO2_mod2, page=1, seWithMean=TRUE)
```

Figure \ref{fig:co2_mod2} shows the fitted smoothers for `CO2_mod2`. The plots of group-specific smoothers indicate that plants differ not only in average log-uptake (which would correspond to each plant having a straight line at different levels for the group-level smoother), but differ slightly in the shape of their functional responses. Figure \ref{fig:co2_mod2_pred} shows how the global and group-specific smoothers combine to predict uptake rates for individual plants.  We see that, unlike in the single global smoother case above, none of the curves deviate from the data systematically.

```{r co2_mod2_ggplot,echo=FALSE, fig.width=6, fig.height=4, out.width="\\linewidth", fig.cap="\\label{fig:co2_mod2_pred}Predicted uptake values (lines) versus observed uptake for each plant, based on model 2."}
CO2_mod2_pred <- predict(CO2_mod2, se.fit=TRUE)
CO2 <- transform(CO2, mod2 = CO2_mod2_pred$fit, mod2_se = CO2_mod2_pred$se.fit)

ggplot(data=CO2, aes(x=conc, y=uptake, group=Plant_uo)) +
  facet_wrap(~Plant_uo) +
  geom_ribbon(aes(ymin=exp(mod2-2*mod2_se),
                  ymax=exp(mod2+2*mod2_se)), alpha=0.25) +
  geom_line(aes(y=exp(mod2))) +
  geom_point() +
  labs(x=expression(CO[2] ~ concentration ~ (mL ~ L^{-1})),
       y=expression(CO[2] ~ uptake ~ (mu*mol ~ m^{-2})))
```

The factor-smoother interaction-based approach mentioned above does not work for higher-dimensional tensor product smoothers. Instead, the group-specific term can be specified with a tensor product of the continuous smoothers and a random effect for the grouping parameter[^fs_note]. e.g.: `y ~ te(x1, x2, bs="tp", m=2) + t2(x1, x2, fac, bs=c("tp","tp","re"), m=2, full=TRUE)`. We illustrate this approach below on the bird migration data. 

[^fs_note]: As mentioned in section II, these terms can be specified either with
`te()` or `t2()` terms. Using `t2` as above (with `full=TRUE`) is essentially a
multivariate equivalent of the factor-smoother interaction; it requires more smooth terms than
`te()`, but can be fit using other mixed effects software such as **lme4**, which
is useful when fitting models with a large number of group levels (see Section
V on computational issues for details).

```{r bird_mod2}
bird_mod2 <- gam(count ~ te(week, latitude, bs=c("cc", "tp"),
                            k=c(10, 10), m=2) +
                   t2(week, latitude, species, bs=c("cc", "tp", "re"),
                      k=c(10, 10, 6), m=2, full=TRUE),
                 data=bird_move, method="REML", family="poisson", 
                 knots = list(week = c(0, 52)))
```

```{r bird_mod2_ggplot, fig.width=8, fig.height=4, echo=FALSE, out.width="\\linewidth", fig.cap="\\label{fig:bird_mod2}a) Predicted migration paths for each species based on \\texttt{bird\\_mod2}, with lighter colors corresponding to higher predicted counts. b) Observed counts versus predictions from \\texttt{bird\\_mod2}."}
bird_move <- transform(bird_move, mod2 = predict(bird_mod2, type="response"))

bird_mod2_indiv <- ggplot(data=bird_move, 
                          aes(x=week, y=latitude, fill=mod2,color=mod2)) +
  geom_tile(size=0.25) +
  facet_wrap(~ species, ncol=6) +
  scale_fill_viridis("Count") +
  scale_color_viridis("Count") +
  scale_x_continuous(expand=c(0, 0), breaks=c(1, 26, 52)) +
  scale_y_continuous(expand=c(0, 0), breaks=c(0, 30, 60)) +
  labs(x = "Week", y = "Latitude") +
  theme(legend.position="right")

bird_mod2_indiv_fit <- ggplot(data=bird_move, aes(x=mod2, y=count)) +
  facet_wrap(~ species, ncol=6) +
  geom_point(alpha=0.1) +
  geom_abline() +
  labs(x="Predicted count (model 2)", y= "Observed count")

plot_grid(bird_mod2_indiv, bird_mod2_indiv_fit, ncol=1, align="vh", axis = "lrtb",
          labels=c("a","b"), rel_heights= c(1,1))
```

Model 2 is able to effectively capture the observed patterns of interspecific
variation in migration behaviour (figure \ref{fig:bird_mod2}a). It shows a much
tighter fit between observed and predicted values, as well as less evidence of
over-dispersion in some species compared to model 1 (figure \ref{fig:bird_mod2}b).

### A single common smoother plus group-level smoothers with differing wiggliness (Model 3)

This model class is very similar to model 2, but we now allow each group-specific smoother to have its own smoothing parameter and hence its own level of wiggliness. This increases the computational cost of the model (as there are more smoothing parameters to estimate), and means that the only information shared between groups is through the global smoother, the common error term, and through the random effect for group-level intercepts (if used). This is useful if different groups differ substantially in how wiggly they are.

Fitting a separate smoother (with its own penalties) can be done in **mgcv** by using the `by` argument in the `s()` and `te()` (and related) functions. Therefore, we can code the formula for this model as:

```r
y ~ s(x, bs="tp") + s(x, by=fac, m=1, bs="tp") + s(fac, bs="re")
```
Note two major differences here from how model 2 was specified:

1.  We explicitly include a random effect for the intercept (the `bs="re"` term), as group-specific intercepts are not incorporated into factor `by` variable smoothers (as would be the case with a factor smoother or a tensor product random effect).
2. We specify `m=1` instead of `m=2` for the groupwise smoothers, which means the marginal TPRS basis for this term will penalize the squared 1st derivative of the function, rather than the second derivative. This, also, reduces co-linearity between the global smoother and the group-specific terms which occasionally leads to high uncertainty around the global smoother (see section V for more details). TPRS with `m=1` have a more restricted null space than m=2 smoothers, so should not be as collinear with the global smoother  [@baayen_autocorrelated_2018; @wieling_investigating_2016]. We have observed that this is much more of an issue when fitting model 3 compared to model 2.


We modify the `CO2` model to follow this approach like so:

```{r mod3_CO2-norun, eval=FALSE}
CO2_mod3 <- gam(log(uptake) ~ s(log(conc), k=5, m=2, bs="tp") +
                  s(log(conc), by=Plant_uo, k=5, m=1, bs="tp") +
                  s(Plant_uo, bs="re", k=12),
                data=CO2, method="REML")
```

```{r mod3_CO2, fig.width=5, fig.height=3, echo=FALSE, fig.cap="\\label{fig:co2_mod3}Functional relationships for the CO2 data estimated for model 3. Top left: the global smoother; Top middle: species-specific random effect intercepts. The remaining plots are a selected subset of the plant-specific smoothers, indicating how the functional response of that plant differs from the global smoother. "}
CO2_mod3 <- gam(log(uptake) ~ s(log(conc), k=5, m=2, bs="tp") +
                  s(log(conc), by= Plant_uo, k=5, m=1, bs="tp") +
                  s(Plant_uo, bs="re", k=12),
                data=CO2, method="REML")

op <- par(mfrow=c(2, 3), mar =c(4, 4, 1, 1))
CO_mod3_ylim = c(-0.3,0.3)
plot(CO2_mod3, scale=0, 
     select=1, 
     ylab="Global smoother", 
     seWithMean=TRUE)
plot(CO2_mod3, 
     scale=0, 
     select=14, 
     ylab="Intercept",
     main=NA)
plot(CO2_mod3, 
     scale=0, 
     select=3,  
     ylab="Plant Qn1", 
     seWithMean=TRUE,
     ylim = CO_mod3_ylim)
plot(CO2_mod3, 
     scale=0, 
     select=5,  
     ylab="Plant Qc1",     
     seWithMean=TRUE,
     ylim = CO_mod3_ylim)
plot(CO2_mod3, 
     scale=0, 
     select=10, 
     ylab="Plant Mn1",     
     seWithMean=TRUE,
     ylim = CO_mod3_ylim)
plot(CO2_mod3, 
     scale=0, 
     select=13, 
     ylab="Plant Mc1",
     seWithMean=TRUE,
     ylim = CO_mod3_ylim)
par(op)
```

Figure \ref{fig:co2_mod3} shows a subsample of the group-specific smoothers from this model. It is apparent from this that some groups (e.g. `Qc1`) have very similar shapes to the global smoother (differing only in intercept), others do differ from the global trend, with higher uptake at low concentrations and lower uptake at higher concentrations (e.g. `Mc1`, `Qn1`), or the reverse pattern (e.g. `Mn1`).

Using model 3 with higher-dimensional data is also straightforward; `by` terms work just as well in tensor-product smoothers as they do with isotropic smoothers. We can see this with our bird model:

```{r mod3_bird}
bird_mod3 <- gam(count ~ species +
                   te(week, latitude, bs=c("cc", "tp"),
                      k=c(10, 10), m=2) +
                   te(week, latitude, by=species, bs= c("cc", "tp"),
                      k=c(10, 10), m=1),
                 data=bird_move, method="REML", family="poisson",
                 knots = list(week = c(0, 52)))
```

As above, we have set (`m=1`) for the latitude marginal effect to  avoid issues of collinearity between the global and groupwise smoother. Note that switching `m=1` to `m=2` does not have any effect on the marginal basis for `week`, where we are using a cyclic smoother instead of a tprs.

The fitted model for `bird_mod3` is visually indistinguishable from `bird_mod2` (figure \ref{fig:bird_mod2}) so we do not illustrate it here. 

### Models without global smoothers (models 4 and 5)

We can modify the above models to exclude the global term (which is generally faster; see section V). When we do not model the global term, we are allowing each factor to be different, though there may be some similarities in the shape of the functions.

#### Model 4:

Model 4 (shared smoothers) is simply model 2 without the global smoother; this type of model takes the form: `y~s(x, fac, bs="fs")` or `y~te(x1, x2, fac, bs=c("tp", "tp", "re")` in `mgcv`. This model assumes all groups have the same smoothness, but that the individual shapes of the smooth terms are not related. Here we just show how to code these models; plotting them works in the same way as for models 1-3 above, the plots for these datasets are very similar to the plots for model 2. This will not always be the case; if in a given study there are very few data points in each grouping level (relative to the strength of the functional relationship of interest), estimates from model 4
will typically be much more variable than from model 2, as there is no way for the model to share information on function shape between grouping levels without the global smoother. See section V on computational issues for more on how to choose between different models.


```{r mod4}
CO2_mod4 <- gam(log(uptake) ~ s(log(conc), Plant_uo, k=5, bs="fs", m=2),
                data=CO2, method="REML")

bird_mod4 <- gam(count ~ t2(week, latitude, species, bs=c("cc", "tp", "re"),
                            k=c(10, 10, 6), m=c(2, 2, 2)),
                 data=bird_move, method="REML", family="poisson",
                 knots = list(week = c(0, 52)))
```

#### Model 5:

Model 5 is simply model 3 without the first term: `y~fac+s(x, by=fac)` or `y~fac+te(x1,x2, by=fac)` (as above, plots are very similar to model 3).

```{r mod5}
CO2_mod5 <- gam(log(uptake) ~ s(log(conc), by=Plant_uo, k=5, bs="tp", m=2) +
                              s(Plant_uo, bs="re", k=12),
                data= CO2, method="REML")


bird_mod5 <- gam(count ~ species + 
                   te(week, latitude, by=species, bs= c("cc", "tp"), 
                      k=c(10, 10), m=2),
                 data=bird_move, method="REML", family="poisson",
                 knots = list(week = c(0, 52)))
```

## Comparing different HGAM specifications

These models can be compared using standard model comparison tools. Model 2 and model 3 will generally be nested in model 1 (depending on how each model is specified) so comparisons using generalized likelihood ratio tests (GLRTs) may be used to test if groupwise smoothers are necessary. However, we do not currently recommend this method. There is not sufficient theory on how accurate parametric p-values are for comparing these models; there is uncertainty about what degrees of freedom to assign to models with varying smoothness, and slightly different model specifications may not result in nested models. (See @wood_generalized_2017 Section 6.12.4 and `?mgcv::anova.gam` for more discussion on using GLRTs to compare GAMs.)

Comparing models based on AIC is a more robust approach to comparing the different model structures. There is well-developed theory of how to include effects of penalization and smoothing parameter uncertainty when estimating the model complexity penalty for AIC [@wood_smoothing_2016]. We demonstrate this approach in Table \ref{tab:AIC_table_kable}. Using AIC, there is strong support for including among-group functional variability for both the CO2 dataset and the `bird_move` dataset (compare models 1 versus models 2-5). For the CO2 dataset (Table \ref{tab:AIC_table_kable}A), there is relatively strong evidence that there is more inter-group variability in smoothness than model 2 allows, and weaker evidence that model 4 or 5 (separate smoothers for all plants) show the best fit. 

For the `bird_move` dataset (Table \ref{tab:AIC_table_kable}B), model 2 (global smoother plus group-level smoothers with a shared penalty) gives the best fit for all models including a global smooth (which is good as we simulated the data from a model with this structure!). However, model 4 (without a global term) still fits this data better than model 2 based on AIC. This highlights an issue with AIC for selecting between models with and without a global smooth: as it is possible to fully recreate the global term by just allowing each group-level smoother to have a similiar shape to one another (that is, the global term is totally concurve with the group-level smoothers; see section V) model selection criteria such as AIC may indicate that the extra parameters required to fit the global smoother are unnecessary. 

```{r AIC_table, echo=FALSE,  fig.width=4, fig.height=6, message=FALSE, warning=FALSE, cache=TRUE}

AIC_table = AIC(CO2_mod1,CO2_mod2, CO2_mod3, CO2_mod4, CO2_mod5,
             bird_mod1, bird_mod2, bird_mod3, bird_mod4, bird_mod5)%>%
  rownames_to_column(var= "Model")%>%
  mutate(data_source = rep(c("CO2","bird_data"), each =5))%>%
  group_by(data_source)%>%
  mutate(deltaAIC = AIC - min(AIC))%>%
  ungroup()%>%
  dplyr::select(-data_source)%>%
  mutate_at(.vars = vars(df,AIC, deltaAIC), 
            .funs = funs(round,.args = list(digits=0)))
```

```{r AIC_table_kable, echo=FALSE,  fig.width=4, fig.height=6, message=FALSE, warning=FALSE, cache=TRUE, purl=FALSE}
kable(AIC_table, format =table_out_format, 
      caption="AIC table comparing model fits for example datasets", 
      booktabs = TRUE)%>% 
  kable_styling(full_width = FALSE)%>%
  group_rows("A. CO2 models", 1,5)%>%
  group_rows("B. bird_move models", 6,10)
```

Given this issue with selecting global terms, we strongly recommend not selecting models based purely on AIC. Instead, model selection should be based on expert subject knowledge about the system, computational time, and most importantly, the inferential goals of the study. Table \ref{tab:AIC_table_kable}A indicates that models 4 and 5 (which do not have a global function) fit the `CO2` data better than models with a global function, and that model 4 fits the `bird_move` data better than model 2. However, it is the shape of the global function that we are actually interested in here, as models 4 and 5 cannot be used to predict the concentration-uptake relationship for plants that are not part of the training set, or the average migration path for birds. The same consideration holds when choosing between model 2 and 3: while model 3 fits the `CO2` data better than model 2 (as measured by AIC), model 2 can be used to simulate functional variation for unobserved group levels, whereas this is not possible within the framework of model 3. The next section works through two examples to show how to choose between different models, and section V discusses these and other model fitting issues in more depth.

It also is important to recognize that AIC, like any function of the data, is a random variable and should be expected to have some sampling error [@forster_aic_2011]. In cases when the goal is to select the model that has the best predictive ability, we recommend holding some fraction of the data out prior to the analysis and comparing how well different models fit that data or using $k$-fold cross validation as a more accurate guide to how well a given model may predict out of sample. 
Predictive accuracy may also be substantially improved by averaging over multiple models [@dormann_model_2018]. 

