
# VI: Examples



***EJP: I think we should aim for 3-4 good examples here, highlighting different
aspects of the model fitting problem. The example I'll be showing is a set of
zooplankton community time series data, where multiple species were tracked
throughout the year for a period of roughtly 20 years. This example can
highlight both testing models 4/5 (for comparing different species' season
cycles) and models 2/3 (for testing for differences between years for a single
species). I think we need at least one example showing how to use these methods
for multivariate regression (e.g. spatial analysis), and potentially an example
showing how to these models work for non-normal data, and for including other
covariates. In all examples, I think we should focus on how to fit each data set, visualize
models, and compare different model fits.***

In this final section, we will go through a few example analyses, to highlight
how to use these models in practice, and to illustrate how to fit, test, and
visualize each model. 


## Example 1: Inter- and intra-specific variation in zooplankton seasonal cycles over time

This first example will demonstrate how to use these models to fit community
data, to show when using a global trend may or may not be justified, and to
illustrate how to use these models to fit seasonal time series. Here, we are
using data from the Wisconsin Department of Natural Resources collected by
Richard Lathrop from a chain of lakes (Mendota, Menona, Kegnonsa, and Waubesa)
in Wisconsin, to study long-term patterns in the seasonal dynamics of
zooplankton. This data consists of roughly bi-weekly samples (during open-water
conditions) of the zooplankton communities, taken from the deepest point of each
lake via vertical tow collected every year from 1976 to 1994 (the collection and
processing of this data is fully described here [*CITE*]). We will use this data
estimate variability in seasonality among species in the community, and between
lakes for the most abundant taxon in the sample (*Daphnia mendotae*). As we are
focusing on seasonal cycles rather than average or maximum abudances, we have
scaled all densities by log-transforming them then scaling by the within year
species- and lake-specific mean and standard deviation (so all species in all
lake-years will have a mean scaled density of zero and standard deviation of
one).

This is what the data looks like:

```{r view_zoo, echo=TRUE, message=FALSE,  cache=TRUE}
zooplankton = read.csv("../data/zooplankton_example.csv")

str(zooplankton)
levels(zooplankton$taxon)
levels(zooplankton$lake)
```

We will split the data into testing and training sets, so we can evaluate how 
well our models fit out of sample. As there are multiple years of data here, 
we will use data from the even years to fit (train) models, and that from the odd years
to test the fit:

```{r zoo_train, echo=TRUE, message=FALSE,  cache=TRUE}
library(mgcv)
zoo_train = subset(zooplankton, year%%2==0) #the modulus (%%) finds the remainder after division by  the right. here we use it to find even numbers

zoo_test = subset(zooplankton, year%%2==1) 
```

Our first exercise here will be to demonstrate how to model community-level
variability in seasonality, by regressing scaled density on day of year, with
species-specific curves. As we are not interested here in average seasonal dynamics,
we will focus on models 4&5[^mean_season]. As this is seasonal data, we will use
cyclic smoothers as the basis for seasonal dynamics. 


[^mean_season]: Here we are focusing on only the most common species
in the data set. If we wanted to estimate the seasonal dynamics for rarer species,
adding a global smooth term might be useful, so we could could borrow information
from the more common species. 

```{r zoo_comm_mod, echo=TRUE, message=FALSE,  cache=TRUE}
zoo_comm_mod4 = gam(density_scaled~s(day, taxon, bs="fs",k=10,xt=list(bs="cc")), 
           data=zoo_train, 
           knots = list(day =c(1,365)), #we need to specify the start and end knots for day
           method = "ML" #We'll use ML as we are comparing models that differ in fixed effects
           )

summary(zoo_comm_mod4)

# as all of the model features except the formula are the same in model 4&5,
# we just use the update function to refit the model with the new formula
zoo_comm_mod5 = update(zoo_comm_mod4,
                       formula = density_scaled~s(day, by=taxon, k=10,bs="cc"))

summary(zoo_comm_mod5)
```

We can see that both models have very similar fits, with an adusted $R^2$ of 0.308 for 
model 4 and 0.31 for model 5.  Model 5 has a somewhat lower AIC (`AIC(zoo_comm_mod4)` = `r round(AIC(zoo_comm_mod4))`, `AIC(zoo_comm_mod5)` = `r round(AIC(zoo_comm_mod5))`),
implying a better overall fit. However, the two models show very similar fit to the data:

```{r zoo_comm_plot, echo=TRUE, message=FALSE, warning=TRUE, cache=TRUE}
library(ggplot2)
library(dplyr)

#Create synthetic data to use to compare predictions
zoo_plot_data = expand.grid(day = 1:365, taxon = factor(levels(zoo_train$taxon)))


zoo_mod4_fit = predict(zoo_comm_mod4, zoo_plot_data, se.fit = T)
zoo_mod5_fit = predict(zoo_comm_mod5, zoo_plot_data, se.fit = T)

zoo_plot_data$mod4_fit = as.numeric(zoo_mod4_fit$fit)
zoo_plot_data$mod5_fit = as.numeric(zoo_mod5_fit$fit)

zoo_plot_data$mod4_se = as.numeric(zoo_mod4_fit$se.fit)
zoo_plot_data$mod5_se = as.numeric(zoo_mod5_fit$se.fit)

zoo_plot = ggplot(zoo_plot_data, aes(x=day))+
  facet_wrap(~taxon, nrow = 2)+
  geom_point(data= zoo_train, aes(y=density_scaled),size=0.1)+
  geom_line(aes(y=mod4_fit))+
  geom_line(aes(y=mod5_fit),color="red")+
  geom_ribbon(aes(ymin = mod4_fit - 2*mod4_se, ymax = mod4_fit + 2*mod4_se), alpha=0.25)+
  geom_ribbon(aes(ymin = mod5_fit - 2*mod5_se, ymax = mod5_fit + 2*mod5_se), alpha=0.25, fill="red")+
  theme_bw()

print(zoo_plot)
```

THe two curves are very close for all species, but the differences in smoothness that resulted 
in model 5 having an higher AIC than model 4 seem to be driven by the low seasonality of *Keratella cochlearis* and *Leptodiaptomus siciloides* relative to the other species. Still, both
models show very similar fits to the training data, model 5 is only slightly better at predicting out of sample fits for *K. cochlearis*, and not at all better for *L. siciloides*: 

```{r zoo_comm_outofsample, echo=TRUE, message=FALSE,  cache=TRUE}
#Getting the out of sample predictions for both models:
zoo_test$mod4 = as.numeric(predict(zoo_comm_mod4,zoo_test))
zoo_test$mod5 = as.numeric(predict(zoo_comm_mod5,zoo_test))

# We'll look at the correlation between fitted and observed values for all species:
zoo_test_summary = zoo_test %>%
  group_by(taxon)%>%
  summarise(mod4_cor = round(cor(density_scaled, mod4),2),
            mod5_cor = round(cor(density_scaled, mod5),2))

print(zoo_test_summary)
```

Now let's look at how to fit inter-lake variability in dynamics for just *Daphnia mendotae*.
Here, we will compare models 1,2, and 3, to determine if a single global function 
is appropriate for all four lakes, or if we can effectively model variation between
lakes with a shared smooth or lake-specific smooths.


```{r zoo_daph_mod, echo=TRUE, message=FALSE,  cache=TRUE}
daphnia_train = subset(zoo_train,taxon=="Daphnia mendotae")
daphnia_test = subset(zoo_test,taxon=="Daphnia mendotae")

zoo_daph_mod1 = gam(density_scaled~s(day, bs="cc",k=10), 
           data=daphnia_train, 
           knots = list(day =c(1,365)), #we need to specify the start and end knots for day
           method = "ML" #We'll use ML as we are comparing models that differ in fixed effects
           )

summary(zoo_daph_mod1)


zoo_daph_mod2 = update(zoo_daph_mod1,
                       formula = density_scaled~s(day, bs="cc",k=10) + s(day,lake, k=10, bs="fs", xt=list(bs="cc")))

summary(zoo_daph_mod2)


zoo_daph_mod3 = update(zoo_daph_mod1,
                       formula = density_scaled~s(day, bs="cc",k=10) + s(day,by=lake, k=10, bs="cc"))

summary(zoo_daph_mod3)
```

The AIC values indicate that both model 2 and 3 are better fits than model 1, but models 2 and 3 have similar fits to one another. There does not seem to be a large amount of inter-lake variability (all three models have similar adjusted $R^2$), and model 3 indicates that only Lake Waubesa deviates substantially from the overall dynamics. The plots for all three models
(model 1 as dashed line, model 2 in black and model 3 in red) show that Medota and Menona 
lakes are very close to the average and to one another for both models (which is unsurprising, 
as they are very closely connected by a short river) but both Kegons and Waubesa show 
evidence of a more pronouced spring bloom and lower winter abundances.  While this is
stronger in Lake Waubesa, model 2 (in black) shows that it is still detectable in
Lake Kegonsa if we do not need to fit a separate penalty for each lake.

```{r zoo_daph_plot, echo=TRUE, message=FALSE, warning=TRUE, cache=TRUE}
library(ggplot2)
library(dplyr)

#Create synthetic data to use to compare predictions
daph_plot_data = expand.grid(day = 1:365, lake = factor(levels(zoo_train$lake)))


daph_mod1_fit = predict(zoo_daph_mod1, daph_plot_data, se.fit = T)
daph_mod2_fit = predict(zoo_daph_mod2, daph_plot_data, se.fit = T)
daph_mod3_fit = predict(zoo_daph_mod3, daph_plot_data, se.fit = T)


daph_plot_data$mod1_fit = as.numeric(daph_mod1_fit$fit)
daph_plot_data$mod2_fit = as.numeric(daph_mod2_fit$fit)
daph_plot_data$mod3_fit = as.numeric(daph_mod3_fit$fit)

daph_plot_data$mod1_se = as.numeric(daph_mod1_fit$se.fit)
daph_plot_data$mod2_se = as.numeric(daph_mod2_fit$se.fit)
daph_plot_data$mod3_se = as.numeric(daph_mod3_fit$se.fit)

daph_plot = ggplot(daph_plot_data, aes(x=day))+
  facet_wrap(~lake, nrow = 2)+
  geom_point(data= daphnia_train, aes(y=density_scaled),size=0.1)+
  geom_line(aes(y=mod1_fit),linetype=2, size=2)+
  geom_line(aes(y=mod2_fit),color="black")+
  geom_line(aes(y=mod3_fit),color="red")+
  geom_ribbon(aes(ymin = mod2_fit - 2*mod2_se, ymax = mod2_fit + 2*mod2_se), alpha=0.25)+
  geom_ribbon(aes(ymin = mod2_fit - 2*mod3_se, ymax = mod2_fit + 2*mod2_se), alpha=0.25, fill="red")+
  theme_bw()

print(daph_plot)
```

In this case, model 2 is able to predict as good or better out of sample as
model 1 or 3, indicating that jointly smoothing the lake together improved model
prediction. However, None of the models did well in terms of predicting Lake
Kegonsa dynamics out of sample (with a correlation of only 0.11 between
predicted and observed densities), indiciating that this model may be be missing
substantial year-to-year variability in *D. mendotae* dynamics:

```{r zoo_daph_outofsample, echo=TRUE, message=FALSE,  cache=TRUE}
#Getting the out of sample predictions for both models:
daphnia_test$mod1 = as.numeric(predict(zoo_daph_mod1,daphnia_test))
daphnia_test$mod2 = as.numeric(predict(zoo_daph_mod2,daphnia_test))
daphnia_test$mod3 = as.numeric(predict(zoo_daph_mod3,daphnia_test))

# We'll look at the correlation between fitted and observed values for all species:
daph_test_summary = daphnia_test %>%
  group_by(lake)%>%
  summarise(mod1_cor = round(cor(density_scaled, mod1),2),
            mod2_cor = round(cor(density_scaled, mod2),2),
            mod3_cor = round(cor(density_scaled, mod3),2))

print(daph_test_summary)
```