---
title: 'Appendix A: Markov Random Fields'
author: "Eric Pedersen"
date: "March 30, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Markov Random Fields

Random effect splines penalize all levels of a grouping variable toward a zero (and thus towards the overall intercept), but otherwise assume the levels of a given group are essentially unrelated; there is no sense in which some levels of a grouping variable should be closer to one another. However, in many ecological problems, we expect the grouping variables to have some form of structure. For instance, we would expect closely related or functionally similar species to respond to ecological drivers such as temperature more similarly to one another that distantly related species. Markov Random field smoothers are one way to incorporate these relationships into our models. 

As with random effects, Markov random field smoothers assign one basis function to each level of a grouping variable. The penalty for the term is then based on a set of connections between groups specified by the user. Groups directly connected to one another will be more strongly penalized toward one another. The resulting penalty matrix $P$ has negative values at  $P_{i,j}=P_{j,i}$ if  group $i$ and $j$ are directly connected, and a positive entry at each diagonal $P_{i,i} \ge -\sum_{j\ne i} P_{i,j}$. As a simple example, imagine a survey with four sites A,B, C, and D, situated along a river in order, so site B is downstream of site A and so on. In this situation we expect estimates for site A to be more similar to B than to C, and more similar to C than to D. The penalty matrix for this case would look like:


$$
\left[ {\begin{array}{cccc}
1  & -1 & 0 & 0 \\
-1 & 2  & -1 & 0 \\
0  & -1 & 2 & -1\\
0& 0 & -1 & 1
 \end{array} } \right]
$$
Interestingly, if you assume that all groups are equally connected to all other groups (so the matrix above would have -1 for all its off-diagonal entries and 3 for all its diagonal entries), you end up with a model that penalizes all groups towards a common mean. This is very similar to how the random effects smoother operates when only modelling group-level means. However, this trick will come in handy later on when we start using interactions between smooth terms to construct hierarchical gams.

The Markov Random Field smoother acts as a smoothing penalty (drawing close groups towards one another). It will also act as a shrinkage penalty as long as one grouping level is dropped (in effect, assigning the global intercept to that level) which is done by default by *mgcv*, and as long as there are no subgroups within the larger group that are entirely unconnected to the rest of the group.


b) The second way of specifying this type of smoother is by using a tensor product of whatever basis you choose for the continuous terms and a Markov Random Field smoother for the factor, with all of the groups in the MRF smoother being fully connected to one another. There is no need to specify a separate term here for the global smooth and group-level smoothers. Instead, this method will directly penalize differences between smooths, simultaneously fitting both the global effect and group-level deviations. e.g.: `y~te(x,fac,bs=c("tp","mrf"),xt=list(penalty=fac_pen_mat))`. This requires a helper function to create the fully connected penalty matrix, which we provide in the supplemental code as `create_connected_pmat`. 


```{r co2_mod2b, echo=TRUE,  fig.width=6, fig.height=3,message=F, warning=F, cache=T}
source("../code/functions.R")
CO2_penmat = create_connected_pmat(CO2$Plant_uo)


CO2_mod2b = gam(log(uptake) ~ te(log(conc),Plant_uo, k=c(5,12),m=2, 
                                 bs=c("tp","mrf"), xt= list(penalty = CO2_penmat)), 
                data= CO2,method="REML")
```

We did not include a call to the plot function for `CO2_mod2b` as there is no default plotting method for a tensor product of a mrf smooth and a continous smoother. 


The two forms of model 2 tend to have similar fits on the same data set. So why not just focus on one or the other? Approach  **a** is useful as it gives a direct estimate of the overall smooth, and it is more straight-forward to incorporate "fixed" effects, where the global term captures large-scale differences between large classes in the data (say between treated and untreated levels in an experiment), and the group-level term models residual among-group functional variance (see [cite Wieling paper here] for a demonstration of this). For instance, in the $CO_2$ uptake analysis above, we could have allowed for average uptake differences between treatments by adding a `by` term: `log(uptake)~s(log(conc), by= Treatment, k=5)+ s(log(conc), Plant_uo, bs="fs",k=5)` (we discuss using the `by=` argument more in the discussion on model 3). Approach **b** does not directly estimate a global smooth, although it can be derived by averaging equivilent basis functions across groups **EJP: should we demonstrate how to do this in paper?**. However, it does require estimating fewer penalty terms, and the general approach can be used to model more complex structural relationships between groups; for instance, if groups are patches in a metapopulation, the fully connected penalty matrix used above could be replaced by a connectivity matrix to determine if interpatch connectivity predicts functional similiarity between patches. In the bird migration case above, we could have used an inverse phylogenetic covariance matrix as the penalty term, to test if closely related species are more similiarly related in migration patterns than distantly related ones. The two models also differ in their computational properties; model 2b is typically faster to estimate, possibly due to issues with concurvity (the nonlinear version of colinearity) between the global and group smooths in model 2a.  We will discuss the computational differences between the two approaches more in section V.
