---
title: "Hierarchical Generalized Additive Models: an introduction with mgcv"
header-includes:
  - \usepackage{lineno}
    \usepackage{placeins}
    \linenumbers
compact-title: FALSE
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
documentclass: "article"
classoption: "12pt"
geometry: left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm,headheight=12pt,letterpaper
bibliography: bibliography.bib
csl: peerj.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE)
library(mgcv)
library(gamm4)
library(tidyr)
library(ggplot2)
library(viridis)
library(cowplot)
library(kableExtra)
library(knitr)
library(tibble)
library(dplyr)

#Set the default theme for ggplot objects to theme_bw()
theme_set(theme_bw())
theme_update(panel.grid = element_blank())
```

# Abstract

In this paper, we discuss an extension to two popular approaches to modelling complex structures in ecological data: the generalized additive model (GAM) and the hierarchical model (HGLM). The hierarchical GAM (HGAM), allows modelling of nonlinear functional relationships between covariates and outcomes
where the shape of the function itself varies between different grouping levels. We describe the theoretical connection between these models,
HGLMs and GAMs, explain how to model different assumptions about the degree of
inter-group variability in functional response, and show how HGAMs can be readily fitted
using existing GAM software, the mgcv package in R. We also discuss computational
and statistical issues with fitting these models, and demonstrate how to fit HGAMs
on example data.

```{r child = '01-intro.Rmd'}
```

```{r child = '02-gams.Rmd'}
```

```{r child = '03-hierarchical_gams.Rmd'}
```

```{r child = '04-examples.Rmd'}
```


```{r child = '05-computational_and_statistical_issues.Rmd'}
```


# Conclusion

HGAMs are a powerful tool to model intergroup variability, and we have attempted
to illustrate some of the range and possibilities that these models are capable
of, how to fit them, and some issues that may arise during
model fitting and testing.
Specifying these models and techniques for fitting them are active areas statistical research, so this paper should be viewed a jumping-off point for these models, rather than an end-point; we refer the reader to the rich literature on GAMs [e.g. @wood_generalized_2017] and functional regression [@ramsay_functional_2005,@kaufman_bayesian_2010, @scheipl_functional_2014] for more on these ideas.



```{r, results='asis',echo=FALSE, eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex')}
#This is just to make sure that the figures occur before the bibliography.
cat('\\FloatBarrier')
```

# Bibliography
