---
title: "Hierarchical Generalized Additive Models in ecology: an introduction with mgcv"
header-includes:
  - \usepackage{lineno}
    \usepackage{placeins}
    \linenumbers
compact-title: FALSE
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
            in_header: author_info.sty
documentclass: "article"
classoption: "12pt"
geometry: left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm,headheight=12pt,letterpaper
bibliography: bibliography.bib
csl: peerj.csl
---

```{r knitr_setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE,fig.path = "../figures/")
```

```{r setup, include=FALSE}
####Setup ####
#all the packages needed for this tutorial are listed here
library(mgcv)
library(MASS)
library(stringr)
library(gamm4)
library(tidyr)
library(ggplot2)
library(viridis)
library(cowplot)
library(kableExtra)
library(docxtools)
library(knitr)
library(tibble)
library(dplyr)

#Set the default theme for ggplot objects to theme_bw()
theme_set(theme_bw())
theme_update(panel.grid = element_blank())
```

```{r setup_tables, include=FALSE, purl=FALSE}
#This code is for configuring latex tables. It is not needed for general script usage.
table_out_format <- ifelse("pdf_document" %in% rmarkdown::all_output_formats("full_document.Rmd"),
                    "latex",
                     ifelse("html_document" %in% rmarkdown::all_output_formats("full_document.Rmd"),
                            "html",
                            NA_character_)
                      )

```

# Abstract

In this paper, we discuss an extension to two popular approaches to modelling complex structures in ecological data: the generalized additive model (GAM) and the hierarchical model (HGLM). The hierarchical GAM (HGAM), allows modelling of nonlinear functional relationships between covariates and outcomes
where the shape of the function itself varies between different grouping levels. We describe the theoretical connection between these models,
HGLMs and GAMs, explain how to model different assumptions about the degree of
inter-group variability in functional response, and show how HGAMs can be readily fitted
using existing GAM software, the mgcv package in R. We also discuss computational
and statistical issues with fitting these models, and demonstrate how to fit HGAMs
on example data. 

```{r child = '01-intro.Rmd'}
```
\FloatBarrier
```{r child = '02-gams.Rmd'}
```
\FloatBarrier
```{r child = '03-hierarchical_gams.Rmd'}
```
\FloatBarrier
```{r child = '04-examples.Rmd'}
```

\FloatBarrier
```{r child = '05-computational_and_statistical_issues.Rmd'}
```


# Conclusion

HGAMs are a powerful tool to model intergroup variability, and we have attempted
to illustrate some of the range and possibilities that these models are capable
of, how to fit them, and some issues that may arise during
model fitting and testing.
Specifying these models and techniques for fitting them are active areas statistical research, so this paper should be viewed as a jumping-off point for these models, rather than an end-point; we refer the reader to the rich literature on GAMs [e.g. @wood_generalized_2017] and functional regression [@ramsay_functional_2005; @kaufman_bayesian_2010; @scheipl_functional_2014] for more on these ideas.

# Acknowledgements

The authors would like to thank Carly Ziter, Tiago Marques, Jake Walsh, Geoff Evans, Paul Regular, and Laura Wheeland for their thoughtful feedback on earlier versions of this manuscript, and the Ecological Society of America for hosting the **mgcv** workshops that this work started from. EJP was funded by National Science and Engineering Research Council of Canada (NSERC) and Fisheries and Oceans Canada. GLS is funded by a Natural Science and Engineering Research Council of Canada (NSERC) Discovery Grant
(RGPIN-2014-04032). DLM was partly funded by OPNAV N45 and the SURTASS LFA Settlement Agreement, managed by the U.S. Navy's Living Marine Resources program under Contract No. N39430-17-C-1982.
NMR was partially funded by the USAID PREDICT-2 program.

All authors contributed to developing the initial idea for this paper, and to writing and editing the manuscript. Author order after the first author was chosen using the code:

```{r eval=FALSE,echo=TRUE, purl=FALSE}
set.seed(11)
sample(c('Miller','Ross','Simpson'))
```

All code used to generate this paper, as well as prior versions of this manuscript, are available at: [github.com/noamross/mixed-effect-gams](https://github.com/noamross/mixed-effect-gams). 

\FloatBarrier


# Bibliography
